---
title: "Altitude data"
author: "Markus Bauer"
date: "`r Sys.Date()`"
output:
   rmarkdown::html_vignette:
      toc: true
vignette: >
  %\VignetteIndexEntry{Get altitude data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette illustrates how you can get altitude data for the wrapper function `check_eunis`. Since the altitude data for Europe is too big for a R package, you have to get it on your own.

## Example

First, we illustrate the altitude data for our example dataset.

```{r libraries, message=FALSE}
library(dplyr)
library(ggplot2)
library(here)
library(sf)
library(terra)
library(tidyterra)
```

### Load the example data

Load the altitude data for the area of our example.

```{r example-data-altitude, warning=FALSE, message=FALSE}
altitude <- terra::rast(
  system.file("extdata", "data_example_altitude.tif", package = "RESY")
)
altitude
```

We load the example plots.

```{r example-data-sites, warning=FALSE, message=FALSE}
data_sites <- readr::read_csv(
  system.file("extdata", "data_example_sites.csv", package = "RESY"),
  show_col_types = FALSE
) |>
  dplyr::select(PlotObservationID, Longitude, Latitude) |>
  sf::st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) |>
  sf::st_transform(terra::crs(altitude))
```

### Map

We show the map with altitude data and the plots

```{r map, message=FALSE}
ggplot() +
  tidyterra::geom_spatraster(data = altitude) +
  tidyterra::scale_fill_whitebox_c(
    palette = "high_relief",
    na.value = "lightblue"
    ) +
  geom_sf(data = data_sites, color = "red", size = 1) +
  labs(fill = "Altitude (m)") +
  theme_minimal()
```

© EuroGeographics 2026, Istituto Geografico Militare (IGM), Italy; [Licence](https://www.mapsforeurope.org/licence)

```{r}
altitude_values <- terra::extract(altitude, terra::vect(data_sites))
data_sites <- data_sites |>
  mutate("Altitude (m)" = altitude_values$eurodem)
data_sites
```

## Crop your own altitude data

Download the elevation data from EuroDEM (European Digital Elevation Model) from <https://www.mapsforeurope.org/datasets/euro-dem> which is from EuroGeographics and the project is cofounded by the European Union.

### Load and adapt EuroDEM data

Save it in your data folder of your R project with your coordinates of your vegetation surveys (sites).

```{r euroddem-data}
# altitude <- terra::rast(
#   here::here("data", "euro-dem-tif", "data", "eurodem.tif")
# )
```

EuroDEM is in arcseconds and we have to rescale it to degrees.

```{r transform-eurodem-data}
# e <- terra::ext(altitude)
# terra::ext(altitude) <- terra::ext(
#   e[1] / 3600, e[2] / 3600, e[3] / 3600, e[4] / 3600
# )
```

The altitude data needs the right coordination system: EPSG:4258 - ETRS89 (European Terrestrial Reference System 1989) in degrees.

```{r assign-epsg-to-eurodem}
# terra::crs(altitude) <- "EPSG:4258"  # geographic ETRS89, degrees
```

Get an overview of the status of the EuroDEM data.

```{r overview-of-eurodem}
# altitude
```

### Load example sites

Now, we can load the sites data and transform it to an sf object and align the CRS to the raster altitude data.

```{r load-data-sites}
# data_sites <- readr::read_csv(
#   here::here("data", "data_example_sites.csv", package = "RESY"),
#   show_col_types = FALSE
# ) |>
#   sf::st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) |>
#   sf::st_transform(terra::crs(altitude))
```

### Crop and save altitude data

We can crop the altitude data to the region of our plots.

```{r crop-data}
# altitude_cropped <- data_sites |>
#   sf::st_buffer(0.5) |> # degrees now, ~50 km
#   sf::st_bbox() |>
#   terra::crop(x = altitude, y = _) |>
#   terra::project("EPSG:4326")
```

Save the smaller file for a latter use.

```{r write-raster}
# terra::writeRaster(
#   altitude_cropped,
#   here::here("data", "data_example_altitude.tif"),
#   overwrite = TRUE
# )
```
