---
title: "Classifying vegetation surveys with RESY"
author: "Florian Jansen, Markus Bauer, Mariasole Calbi & Gilles Colling"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Classifying vegetation surveys with RESY}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

RESY implements the **Expert System (ESy)** framework for vegetation
classification (Bruelheide [1997](https://doi.org/10.1007/BF02803883); Chytrý et al. [2020](https://doi.org/10.1111/avsc.12519)).
An ESy file encodes three sections:

1. **Taxonomy** – species aggregations and synonyms  
2. **Taxon groups** – species assemblages referenced in classification conditions  
3. **Vegetation types** – logical formulas that combine group thresholds into
   named habitat types

The package can work with any ESy-compliant classification. This vignette uses
two bundled systems:

- **EUNIS** (European Nature Information System habitat types;
  [FloraVeg.EU](https://floraveg.eu/habitat/))
- **Apennine-test** (a simplified test system for the Italian Apennines)

For EUNIS specifically, many formulas also require geographic columns
(`Ecoreg`, `Country`, `Coast_EEA`, `Dunes_Bohn`) in a sites header table.
Those preparation steps are covered in `vignette("EUNIS")`.

```{r libraries, echo=TRUE, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(RESY)
```

---

## 1. Load and prepare species data

`resy_classify()` expects the observation table to have three columns:

| Column | Description |
|---|---|
| `PlotObservationID` | Plot identifier (auto-detected) |
| `TaxonName` | Species name |
| `Cover_Perc` | Cover value (percent or Braun-Blanquet scale) |

```{r load-species, message=FALSE}
data_species <- read_csv(
  system.file("extdata", "data_example_species.csv", package = "RESY"),
  show_col_types = FALSE
) |>
  rename(TaxonName = species, Cover_Perc = cover)

glimpse(data_species)
```

A minimal **header** table (one row per plot) is required even when no
geographic conditions are used. It must contain the same plot ID column as
the observation table.

```{r header}
header <- data_species |>
  distinct(PlotObservationID) |>
  as.data.frame()
```

---

## 2. Load expert systems

`resy_available_classifications()` lists the schemes and versions bundled with the package.

```{r available}
resy_available_classifications() |> select(scheme, version)
```

To add a custom expert system, use `resy_add_classification()`.


`resy_load_expert()` parses a bundled classification into a
`resy_parsed_expert` object, used here for taxonomy checking.
`resy_classify()` loads the expert internally from `scheme`/`version`.

```{r load-experts}
parsed_apennine <- resy_load_expert(scheme = "Apennine-test")
```



## 3. Check taxonomy

`resy_check_taxonomy()` maps observation names against Section 1 of the expert
system (canonical names + synonyms). This is classification-specific: a name
may resolve in one system but not another.


```{r taxonomy-apennine}
tax_apennine <- resy_check_taxonomy(
  obs    = data_species,
  parsed = parsed_apennine,
  col    = "TaxonName"
)
cat("Apennine-test — matched:", sum(tax_apennine$matched),
    "/ unmatched:", sum(!tax_apennine$matched), "\n")
```


## 4. Classify

`resy_classify()` evaluates all membership formulas and returns a
`resy_result` object. Pass `scheme` to select the bundled expert system; it
will be loaded automatically.


```{r classify-apennine, message=FALSE}
res_apennine <- resy_classify(
  obs    = data_species,
  header = header,
  scheme = "Apennine-test"
)
```

---

## 5. Inspect results

### Print classification hierarchy

```{r tree}
tree_filled <- resy_expert_tree(parsed_apennine, fill = TRUE)
print(tree_filled)
```

To browse the type hierarchy of a loaded system, use `resy_view_expert()`.


### Vegetation type details

`resy_eval_type()` shows conditiosn of a type is evaluated.

```{r eval-type}
resy_eval_type(res_apennine, t = "FB")
```


### Long table of candidates for all plots

`resy_candidates()` extracts ranked classification results. Use `top_n` to
limit the number of types returned per plot.

```{r candidates-apennine}
cand_apennine <- resy_candidates(res_apennine, top_n = 3)
head(cand_apennine)

```

### Plot-level details

`resy_eval_plot()` prints the full evidence for one plot: which species
matched, which group conditions fired, and which vegetation-type formulas
evaluated to `TRUE`.

```{r eval-plot}
# Replace "AN57" with a PlotObservationID present in your data
resy_eval_plot(res_apennine, p = "AN57")

```




## Next steps

- For EUNIS with geographic enrichment (ecoregion, country, coast), see
  `vignette("EUNIS")`.
- To validate an ESy text file before importing, use `resy_validate_esy()`.

