---
title: "RESY taxonomy"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output:
   rmarkdown::html_vignette:
      toc: true
vignette: >
  %\VignetteIndexEntry{RESY taxonomy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette illustrates how to harmonise the species names of a vegetation
survey before it is classified with `resy_classify()`. RESY implements the
**Expert System (ESy)** framework (Chytrý et al.
[2020](https://doi.org/10.1111/avsc.12519); Bruelheide et al.
[2021](https://doi.org/10.1111/avsc.12562)), and `resy_classify()` matches the
species names it is given against the vocabulary of the expert system. That works
directly for data from the European Vegetation Archive. Plot data named under a
general backbone (GBIF, World Flora Online, POWO, Euro+Med, a national checklist)
often uses different strings for the same taxon: an accepted-name synonym, an
author citation, or a spelling variant. Those names have to be harmonised to the
expert vocabulary first. RESY treats this as a separate step, so you can check
the taxonomic decisions before they affect a classification.

1. **Resolve** the species names to the vocabulary of the expert system with
   `resy_resolve_taxa()`.
2. **Inspect** the result with `resy_summarize_taxa()`.
3. **Classify** the resolved data with `resy_classify()`.

## Example

The following R packages are required to run the example of this vignette:

```{r setup, message=FALSE}
library(RESY)
library(ggplot2)
```

## The shipped synonym table

RESY ships a synonym table that maps alternate species names to the canonical
name used by the EUNIS-ESy expert system. It pools synonymy from six backbones
(Euro+Med, WFO, GBIF, Catalogue of Life, ITIS, NCBI), so a name coming from any
of them can resolve without you having to say which backbone it came from.

```{r shipped-table}
syn <- resy_read_synonyms()
nrow(syn)
head(syn)
```

Each row is one alternate spelling (`synonym`), the canonical ESy name it
resolves to (`esy_canonical`), and the backbone(s) that support the pairing
(`source`).

A pair is listed under every backbone that supports it, so the counts per
backbone add up to more than the number of rows.

```{r support-table}
per_backbone <- table(unlist(strsplit(syn$source, ";", fixed = TRUE)))
support <- data.frame(
  backbone = names(per_backbone),
  pairs    = as.integer(per_backbone),
  share    = round(100 * as.integer(per_backbone) / nrow(syn), 1)
)
support <- support[order(-support$pairs), ]

knitr::kable(
  support,
  row.names   = FALSE,
  col.names   = c("Backbone", "Synonym pairs", "% of table"),
  format.args = list(big.mark = ",")
)
```

Most pairs are supported by more than one backbone:

```{r support-plot, fig.width=6, fig.height=3}
per_pair <- table(lengths(strsplit(syn$source, ";", fixed = TRUE)))
support_n <- data.frame(
  n_backbones = names(per_pair),
  pairs       = as.integer(per_pair)
)

ggplot(support_n, aes(x = n_backbones, y = pairs)) +
  geom_col() +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
  labs(x = "Number of backbones supporting a pair", y = "Synonym pairs") +
  theme_classic()
```

## Resolve names

`resy_resolve_taxa()` maps a column of names to the canonical vocabulary. A name
already equal to a canonical name resolves to itself (`"exact"`); otherwise it is
looked up in the synonym table (`"synonym"`). A name that matches neither is
matched again with its author citation removed by `resy_clean_names()`
(`"cleaned_exact"`, `"cleaned_synonym"`); the name in your data is left as it
is. A name that still matches nothing stays `NA` and is flagged `"unresolved"`.
Nothing is ever replaced with a best guess.

```{r resolve}
plots <- data.frame(
  TaxonName = c(
    "Hibiscus praecox",        # a synonym in the shipped table
    "Abelmoschus longifolius",  # another synonym
    "Fagus sylvatica",          # already a canonical name
    "Anemone nemorosa L.",      # canonical name with an author citation
    "Not a real species 123"    # resolves to nothing
  ),
  stringsAsFactors = FALSE
)

resolved <- resy_resolve_taxa(plots, species_col = "TaxonName")
resolved
```

`resy_summarize_taxa()` reports how many names resolved, the breakdown by
confidence, and the distinct names that stayed unresolved.

```{r summarise}
resy_summarize_taxa(resolved, species_col = "TaxonName")
```

Unresolved names can be fixed by correcting a spelling, adding a row to your own
crosswalk, or leaving them out.

## Bring your own crosswalk

`resy_resolve_taxa()` also accepts your own crosswalk, with three columns:
`synonym`, `esy_canonical`, and `source`. A small example ships with the package.

```{r own-csv}
cw_path <- system.file("extdata", "example_crosswalk.csv", package = "RESY")
crosswalk <- resy_read_synonyms(cw_path)
crosswalk
```

Pass it to `synonyms=`, and give the target names to `canonical=` so resolution
uses your table alone:

```{r own-resolve}
my_plots <- data.frame(
  TaxonName = c("Fagus silvatica", "Abies pectinata", "Quercus robur"),
  stringsAsFactors = FALSE
)

resy_resolve_taxa(
  my_plots,
  species_col = "TaxonName",
  synonyms    = crosswalk,
  canonical   = unique(crosswalk$esy_canonical)
)
```

If your crosswalk lives in an Excel sheet, read it yourself and pass the data
frame. RESY validates the columns and uses it directly, so it needs no Excel
package as a dependency:

```{r excel, eval=FALSE}
crosswalk <- readxl::read_excel("my_crosswalk.xlsx")
resy_resolve_taxa(my_plots, synonyms = crosswalk,
                  canonical = unique(crosswalk$esy_canonical))
```

A synonym that maps to more than one canonical name is dropped, so an ambiguous
table cannot produce a silent substitution.

## Classify the resolved data

Once names are in the expert vocabulary, classify as usual. Resolve them against
the vocabulary of the expert system you classify with (here `Apennine-test`),
write the resolved canonical name back to the name column, keeping the original
where a name did not resolve, and pass the result to `resy_classify()`.

```{r classify, message=FALSE}
library(readr)

species <- read_csv(
  system.file("extdata", "data_example_species.csv", package = "RESY"),
  show_col_types = FALSE
)
species <- head(species, 200)   # keep the example quick

obs <- data.frame(
  PlotObservationID = species$PlotObservationID,
  TaxonName         = species$species,
  Cover_Perc        = species$cover,
  stringsAsFactors  = FALSE
)
expert <- resy_load_expert(scheme = "Apennine-test")
expert_vocab <- unique(c(names(expert$aggs), unlist(expert$aggs, use.names = FALSE)))

resolved_obs <- resy_resolve_taxa(
  obs, species_col = "TaxonName", canonical = expert_vocab
)
table(resolved_obs$taxon_confidence)

resolved_obs$TaxonName <- ifelse(
  is.na(resolved_obs$canonical), resolved_obs$TaxonName, resolved_obs$canonical
)
obs <- resolved_obs[c("PlotObservationID", "TaxonName", "Cover_Perc")]
header <- as.data.frame(unique(obs["PlotObservationID"]))

res <- resy_classify(obs, header, scheme = "Apennine-test")
head(res$result.classification)
```

For a quick pass, `resy_classify(..., resolve_taxa = TRUE)` runs the resolution
against the expert vocabulary and then classifies in a single call, and reports
what it did in `res$taxon_resolution`. The two-step route above is recommended,
since it lets you check the unresolved names before they reach the classifier.

## Building a crosswalk with taxify

To assemble a crosswalk from offline backbone tables (no API calls), the
[taxify](https://github.com/gcol33/taxify) package matches names against
Euro+Med, WFO, GBIF, Catalogue of Life, ITIS, and NCBI and records the supporting
backbone for each match. A table with the columns `synonym`, `esy_canonical`, and
`source` built from those matches can be passed to `resy_resolve_taxa()` as
`synonyms=`.

## References

Bruelheide H, Tichý L, Chytrý M, Jansen F ([2021](https://doi.org/10.1111/avsc.12562)) Implementing the formal language of the vegetation classification expert systems (ESy) in the statistical computing environment R. &ndash; _Applied Vegetation Science_ 24, e12562 [https://doi.org/10.1111/avsc.12562](https://doi.org/10.1111/avsc.12562)

Chytrý M, Tichý L, Hennekens SM et al. ([2020](https://doi.org/10.1111/avsc.12519)) EUNIS Habitat Classification: expert system, characteristic species combinations and distribution maps of European habitats. &ndash; _Applied Vegetation Science_ 23, 648&ndash;675. [https://doi.org/10.1111/avsc.12519](https://doi.org/10.1111/avsc.12519)
