---
title: "Harmonising taxonomy before classification"
output:
   rmarkdown::html_vignette:
      toc: true
vignette: >
  %\VignetteIndexEntry{Harmonising taxonomy before classification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(RESY)
```

## Why a separate step

`resy_classify()` is taxonomy-agnostic: it classifies the species names it is
given, matching them against the vocabulary of the expert system. It works best
when the names in your plot data already match that vocabulary, which is the case
for data drawn from the European Vegetation Archive.

Plot data named under a general backbone (GBIF, World Flora Online, POWO,
Euro+Med, a national checklist) will often use different strings for the same
taxon: an accepted-name synonym, an author citation, or a spelling variant. Those
names have to be harmonised to the expert vocabulary first. RESY keeps that as an
explicit step rather than hiding it inside the classifier, so the taxonomic
decisions stay visible and you can check them before they affect a
classification.

The workflow is: resolve names, inspect the result, then classify.

## The shipped synonym table

RESY ships a synonym table that maps alternate species names to the canonical
name used by the EUNIS-ESy expert system. It pools synonymy from six backbones
(Euro+Med, WFO, GBIF, Catalogue of Life, ITIS, NCBI), so a name coming from any
of them can resolve without you having to say which backbone it came from.

```{r shipped-table}
syn <- resy_read_synonyms()
nrow(syn)
head(syn)
```

Each row is one alternate spelling (`synonym`), the canonical ESy name it
resolves to (`esy_canonical`), and the backbone(s) that support the pairing
(`source`).

## Resolve names

`resy_resolve_taxa()` maps a column of names to the canonical vocabulary. A name
already equal to a canonical name resolves to itself (`"exact"`); otherwise it is
looked up in the synonym table (`"synonym"`); a name matching neither stays `NA`
and is flagged `"unresolved"`. Nothing is ever replaced with a best guess.

```{r resolve}
plots <- data.frame(
  TaxonName = c(
    "Hibiscus praecox",        # a synonym in the shipped table
    "Abelmoschus longifolius",  # another synonym
    "Fagus sylvatica",          # already a canonical name
    "Not a real species 123"    # resolves to nothing
  ),
  stringsAsFactors = FALSE
)

resolved <- resy_resolve_taxa(plots, species_col = "TaxonName")
resolved
```

`resy_summarize_taxa()` reports how many names resolved, the breakdown by
confidence, and the distinct names that stayed unresolved.

```{r summarise}
resy_summarize_taxa(resolved, species_col = "TaxonName")
```

The unresolved names are yours to handle: correct a spelling, add a row to your
own crosswalk, or leave them out. The package will not decide for you.

## Bring your own crosswalk

You are not tied to the shipped table. `resy_resolve_taxa()` accepts any
crosswalk with three columns: `synonym`, `esy_canonical`, and `source`. A small
example ships with the package.

```{r own-csv}
cw_path <- system.file("extdata", "example_crosswalk.csv", package = "RESY")
crosswalk <- resy_read_synonyms(cw_path)
crosswalk
```

Pass it to `synonyms=`, and give the target names to `canonical=` so resolution
uses your table alone:

```{r own-resolve}
my_plots <- data.frame(
  TaxonName = c("Fagus silvatica", "Abies pectinata", "Quercus robur"),
  stringsAsFactors = FALSE
)

resy_resolve_taxa(
  my_plots,
  species_col = "TaxonName",
  synonyms    = crosswalk,
  canonical   = unique(crosswalk$esy_canonical)
)
```

If your crosswalk lives in an Excel sheet, read it yourself and pass the data
frame. RESY validates the columns and uses it directly, so no Excel package is
added as a dependency and the file stays in a format you can review:

```{r excel, eval=FALSE}
crosswalk <- readxl::read_excel("my_crosswalk.xlsx")
resy_resolve_taxa(my_plots, synonyms = crosswalk,
                  canonical = unique(crosswalk$esy_canonical))
```

A synonym that maps to more than one canonical name is dropped rather than
resolved to an arbitrary pick, so an ambiguous table never produces a silent
substitution.

## Classify the resolved data

Once names are in the expert vocabulary, classify as usual. Write the resolved
canonical name back to the name column, keeping the original where a name did not
resolve, and pass it to `resy_classify()`.

```{r classify, message=FALSE}
library(readr)

species <- read_csv(
  system.file("extdata", "data_example_species.csv", package = "RESY"),
  show_col_types = FALSE
)
species <- head(species, 200)   # keep the example quick

obs <- data.frame(
  PlotObservationID = species$PlotObservationID,
  TaxonName         = species$species,
  Cover_Perc        = species$cover,
  stringsAsFactors  = FALSE
)
header <- as.data.frame(unique(obs["PlotObservationID"]))

res <- resy_classify(obs, header, scheme = "Apennine-test")
head(res$result.classification)
```

For a quick pass, `resy_classify(..., resolve_taxa = TRUE)` runs the resolution
against the expert vocabulary and then classifies in a single call, and reports
what it did in `res$taxon_resolution`. Doing it in two steps, as above, is the
recommended route: it keeps the resolution result in your hands so you can check
the unresolved names before they reach the classifier.

## Building a crosswalk with taxify

To assemble a crosswalk from offline backbone tables (no API calls), the
[taxify](https://github.com/gcol33/taxify) package matches names against
Euro+Med, WFO, GBIF, Catalogue of Life, ITIS, and NCBI and records the supporting
backbone for each match. Its output maps to the `synonym` / `esy_canonical` /
`source` schema `resy_resolve_taxa()` expects.
